---
title: "Maine Transportation Data"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
runtime: shiny

---

```{r setup, include=FALSE, cache=T}
library(flexdashboard)
library(shiny)
library(leaflet)
library(tidyverse)
library(tidycensus)
library(tidyverse)
library(treemapify)
library(sf)
library(DT)
library(stringr)
library(mapproj)
# library(ggcounty)
# census_api_key("7dbb06ee22682429da281391496c9628376b9358", install=T)
me <- read.csv("me.csv")[,-1]
maine <- readRDS("maine.RDS")
me_means_transport <-
  me %>% filter(
    concept == "MEANS OF TRANSPORTATION TO WORK FOR WORKPLACE GEOGRAPHY" &
      label %in% c(
        "Estimate!!Total",
        "Estimate!!Total!!Car, truck, or van",
        "Estimate!!Total!!Public transportation (excluding taxicab)",
        "Estimate!!Total!!Bicycle",
        "Estimate!!Total!!Walked",
        "Estimate!!Total!!Worked at home"
      )
  ) %>% mutate(
    mode = str_remove(label, "Estimate!!Total!!"),
    NAME = str_remove(NAME, " County, Maine"))
    
me_traveltime <-
  me %>% filter(concept == "TRAVEL TIME TO WORK FOR WORKPLACE GEOGRAPHY")  %>% mutate(time = str_remove(label, "Estimate!!Total!!"))
    
me_arrivetime <-
  me %>% filter(concept == "TIME ARRIVING AT WORK FROM HOME FOR WORKPLACE GEOGRAPHY") %>% mutate(time = str_remove(label, "Estimate!!Total!!"))
```

# About 


**What is the purpose of this dashboard?**

This dashboard allows for the exploration of transportation data in Maine counties. The data come from the 2017  American Community Survey on the county level and can be obtained using the [tidycensus API](https://walker-data.com/tidycensus/). 


# Data


-------------------------------------


The data used in this dashboard are from the [2017 American Community Survey](https://www.census.gov/programs-surveys/acs) and were obtained using the [tidycensus API](https://walker-data.com/tidycensus/). 



*The American Community Survey (ACS) helps local officials, community leaders, and businesses understand the changes taking place in their communities. It is the premier source for detailed population and housing information about our nation.* 

***

```{r, echo=FALSE}
# Create the DT table first
DT_table <- me %>%
  dplyr::select(NAME, concept, estimate, moe, label) %>%
  rename(c("County" = NAME, "Margin of Error" = moe, "Variable" = concept, "Label" = label, "Estimate" = estimate))
# Instead of depending on the st_jitter algorithm to generate random placement, a custom function placing the points side by side at a set distance could be used to make points occuring at the same location appear neatly apart.
```

```{r, echo=FALSE}
DT::renderDataTable({
  DT::datatable(DT_table,
                caption = htmltools::tags$caption(
                  style = 'caption-side: top; text-align: Left;'),
                options = list(autoWidth = TRUE,
                               pageLength = 10,
                               scroller = TRUE,
                               scrollY = '450px'))
})
```


# Commuting: Foot and Bicycle Maps

Column {data-width=500}
-------------------------------------
### By Bicycle

```{r}

gg <- maine$gg
# add a new geom with our population (choropleth)

me_small <- me %>% mutate(
  cname = word(NAME, 1)) %>% 
  filter(concept == "MEANS OF TRANSPORTATION TO WORK FOR WORKPLACE GEOGRAPHY") %>% 
  group_by(GEOID, cname) %>% summarize(
    total = estimate[label == "Estimate!!Total"],
    bike = (estimate[label == "Estimate!!Total!!Bicycle"]/total)*100
  )
countynames <- aggregate(cbind(long, lat) ~ id, data=maine$map, 
                    FUN=function(x)mean(range(x)))
gg + geom_map(data=me_small, map=maine$map,
                    aes(map_id=cname, fill=bike), 
                    color="white", size=0.125)+scale_fill_viridis_c(name = "% Cyclists of Total Commuters", limits = c(0, .8))+geom_label(data=countynames, aes(long, lat, label = id), size=2)+labs(title = "Bike Commuting in Maine Counties")

```

Column {data-width=500}
-------------------------------------
### By Foot

```{r}

me_small <- me %>% mutate(
  cname = word(NAME, 1)) %>% 
  filter(concept == "MEANS OF TRANSPORTATION TO WORK FOR WORKPLACE GEOGRAPHY") %>% 
  group_by(GEOID, cname) %>% summarize(
    total = estimate[label == "Estimate!!Total"],
    walk = (estimate[label == "Estimate!!Total!!Walked"]/total)*100
  )

gg + geom_map(data=me_small, map=maine$map,
                    aes(map_id=cname, fill=walk), 
                    color="white", size=0.125)+scale_fill_viridis_c(name = "% Walkers of Total Commuters")+geom_label(data=countynames, aes(long, lat, label = id), size=2)+labs(title = "Commuting by Foot in Maine Counties")

```


# Driving and Public Transportation 
Column 
-------------------------------------
    
### Public Transportation
    
```{r}

gg <- maine$gg
# add a new geom with our population (choropleth)

me_small <- me %>% mutate(
  cname = word(NAME, 1)) %>% 
  filter(concept == "MEANS OF TRANSPORTATION TO WORK FOR WORKPLACE GEOGRAPHY") %>% 
  group_by(GEOID, cname) %>% summarize(
    total = estimate[label == "Estimate!!Total"],
    car = (estimate[label == "Estimate!!Total!!Public transportation (excluding taxicab)"]/total)*100
  )
countynames <- aggregate(cbind(long, lat) ~ id, data=maine$map, 
                    FUN=function(x)mean(range(x)))
gg + geom_map(data=me_small, map=maine$map,
                    aes(map_id=cname, fill=car), 
                    color="white", size=0.125)+scale_fill_viridis_c(name = "% of Total Commuters")+geom_label(data=countynames, aes(long, lat, label = id), size=2)+labs(title = "Commute by Public Transportation in Maine Counties")
```
   
Column {.tabset}
-------------------------------------
   
### Car, Truck, or Van Carpool 

```{r}

me_small <- me %>% mutate(
  cname = word(NAME, 1)) %>% 
  filter(concept == "MEANS OF TRANSPORTATION TO WORK FOR WORKPLACE GEOGRAPHY") %>% 
  group_by(GEOID, cname) %>% summarize(
    total = estimate[label == "Estimate!!Total"],
    carpool = (estimate[label == "Estimate!!Total!!Car, truck, or van!!Carpooled"]/total)*100
  )
gg + geom_map(data=me_small, map=maine$map,
                    aes(map_id=cname, fill=carpool), 
                    color="white", size=0.125)+scale_fill_viridis_c(name = "% of Total Commuters")+geom_label(data=countynames, aes(long, lat, label = id), size=2)+labs(title = "Commute by Carpool in Maine Counties")
```   
 
### Car, Truck, or Van Alone 
    
```{r}

gg <- maine$gg
# add a new geom with our population (choropleth)

me_small <- me %>% mutate(
  cname = word(NAME, 1)) %>% 
  filter(concept == "MEANS OF TRANSPORTATION TO WORK FOR WORKPLACE GEOGRAPHY") %>% 
  group_by(GEOID, cname) %>% summarize(
    total = estimate[label == "Estimate!!Total"],
    car = (estimate[label == "Estimate!!Total!!Car, truck, or van!!Drove alone"]/total)*100
  )
countynames <- aggregate(cbind(long, lat) ~ id, data=maine$map, 
                    FUN=function(x)mean(range(x)))
gg + geom_map(data=me_small, map=maine$map,
                    aes(map_id=cname, fill=car), 
                    color="white", size=0.125)+scale_fill_viridis_c(name = "% of Total Commuters")+geom_label(data=countynames, aes(long, lat, label = id), size=2)+labs(title = "Commute by Car, Truck, or Van Alone in Maine Counties")
```



# Compare Commuting by County 


Column {.sidebar}
-----------------------------------------------------------------------

Commuting data in Maine by county

```{r}

checkboxGroupInput("counties", label = "Choose counties",
            choices = c("Androscoggin" = "Androscoggin County, Maine", "Aroostook" = "Aroostook County, Maine",    "Cumberland" = "Cumberland County, Maine" ,  "Franklin" ="Franklin County, Maine",    "Hancock"="Hancock County, Maine",      "Kennebec"="Kennebec County, Maine",     "Knox"="Knox County, Maine",  "Lincoln"="Lincoln County, Maine",      "Oxford" ="Oxford County, Maine",      "Penobscot" ="Penobscot County, Maine",   "Piscataquis"= "Piscatiquis County, Maine",  "Sagadahoc" = "Sagadahoc County, Maine" ,   "Somerset"="Somerset County, Maine"  ,   "Waldo" = "Waldo County, Maine",       "Washington" = "Washington County, Maine"  , "York"="York County, Maine" ), selected = c("Androscoggin County, Maine", "Aroostook County, Maine"))



```

Row
-----------------------------------------------------------------------

### Percentage of Commuter Types  


```{r}

renderPlot({
  
    df <- me %>% filter(NAME %in% input$counties & concept == "MEANS OF TRANSPORTATION TO WORK FOR WORKPLACE GEOGRAPHY") %>% group_by(NAME) %>% summarize(
    total = estimate[label == "Estimate!!Total"],
    Biked = (estimate[label == "Estimate!!Total!!Bicycle"]/total)*100,
    Walked= (estimate[label == "Estimate!!Total!!Walked"]/total)*100,
    Carpooled = (estimate[label == "Estimate!!Total!!Car, truck, or van!!Carpooled"]/total)*100,
    `Drove Alone` = (estimate[label == "Estimate!!Total!!Car, truck, or van!!Drove alone"]/total)*100,
    `Public Transport`= (estimate[label == "Estimate!!Total!!Public transportation (excluding taxicab)"]/total)*100,
    `Worked at Home` = (estimate[label == "Estimate!!Total!!Worked at home"]/total)*100
  ) %>% pivot_longer(cols =-c(NAME, total))
  ggplot(df, aes(x = NAME, y=  value, fill= as.factor(NAME)))+facet_wrap(.~name, scales="free") +geom_bar(position = "dodge",
                                           stat = "identity", col = "black")+scale_fill_viridis_d(name = "")+theme_minimal()+
    theme(legend.position="bottom")+labs(y= "Percent", x = "")+theme(axis.text.x=element_blank())
  

  
})
```

Row 
---------------------------
### Absolute Number of Commuters 

```{r}
renderPlot({
df <- me %>% filter(NAME %in% input$counties & concept == "MEANS OF TRANSPORTATION TO WORK FOR WORKPLACE GEOGRAPHY") %>% group_by(NAME) %>% summarize(
    total = estimate[label == "Estimate!!Total"],
    Biked = (estimate[label == "Estimate!!Total!!Bicycle"]),
    Walked= (estimate[label == "Estimate!!Total!!Walked"]),
    Carpooled = (estimate[label == "Estimate!!Total!!Car, truck, or van!!Carpooled"]),
    `Drove Alone` = (estimate[label == "Estimate!!Total!!Car, truck, or van!!Drove alone"]),
    `Public Transport`= (estimate[label == "Estimate!!Total!!Public transportation (excluding taxicab)"]),
    `Worked at Home` = (estimate[label == "Estimate!!Total!!Worked at home"])
  ) %>% pivot_longer(cols =-c(NAME, total))
  ggplot(df, aes(x = NAME, y=  value, fill= as.factor(NAME)))+facet_wrap(.~name, scales="free")+geom_bar(position = "dodge",
                                           stat = "identity", col = "black")+scale_fill_viridis_d(name = "")+theme_minimal()+
    theme(legend.position="bottom")+labs(y= "Count", x = "")+theme(axis.text.x=element_blank())
})
```



# Compare Arrival Times and Length of Commute by County


Column {.sidebar}
-----------------------------------------------------------------------



```{r}
checkboxGroupInput("countiesradio", label = "Choose counties",
            choices = c("Androscoggin" = "Androscoggin County, Maine", "Aroostook" = "Aroostook County, Maine",    "Cumberland" = "Cumberland County, Maine" ,  "Franklin" ="Franklin County, Maine",    "Hancock"="Hancock County, Maine",      "Kennebec"="Kennebec County, Maine",     "Knox"="Knox County, Maine",  "Lincoln"="Lincoln County, Maine",      "Oxford" ="Oxford County, Maine",      "Penobscot" ="Penobscot County, Maine",   "Piscataquis"= "Piscatiquis County, Maine",  "Sagadahoc" = "Sagadahoc County, Maine" ,   "Somerset"="Somerset County, Maine"  ,   "Waldo" = "Waldo County, Maine",       "Washington" = "Washington County, Maine"  , "York"="York County, Maine" ), selected =c( "Androscoggin County, Maine", "Kennebec County, Maine"))


```

Row {.tabset}
-----------------------------------------------------------------------

### Arrival Time at Work 


```{r}

renderPlot({
  me_arrivetime %>% filter(NAME %in% input$countiesradio) %>% group_by(NAME) %>%
  mutate(
    pct = estimate / (estimate[label == "Estimate!!Total"]),
    NAME = str_remove(NAME, " County, Maine"),
    time = factor(
      time,
      levels = c(
        "12 00 a.m. to 4 59 a.m.",
        "5 00 a.m. to 5 29 a.m.",
        "5 30 a.m. to 5 59 a.m.",
        "6 00 a.m. to 6 29 a.m." ,
        "6 30 a.m. to 6 59 a.m."  ,
        "7 00 a.m. to 7 29 a.m." ,
        "7 30 a.m. to 7 59 a.m."  ,
        "8 00 a.m. to 8 29 a.m.",
        "8 30 a.m. to 8 59 a.m."  ,
        "9 00 a.m. to 9 59 a.m." ,
        "10 00 a.m. to 10 59 a.m.",
        "11 00 a.m. to 11 59 a.m." ,
        "12 00 p.m. to 3 59 p.m." ,
        "4 00 p.m. to 11 59 p.m."
      ),
      labels = c(
        "12-4AM",
        "5-5:30AM",
        "5:30-6AM",
        "6-6:30AM" ,
        "6:30-7AM"  ,
        "7-7:30AM" ,
        "7:30-8AM"  ,
        "8-8:30AM",
        "8:30-9AM"  ,
        "9-10AM" ,
        "10-11AM",
        "11-12AM" ,
        "12-4PM" ,
        "4-12PM"
      ),
      ordered = T
    )
  ) %>%
  filter(label != "Estimate!!Total") %>%
  ggplot(aes(
    area = pct,
    fill = time,
    label = paste(time, ": ", round(pct*100, 0), "%", sep="")
  )) + geom_treemap() + geom_treemap_text(colour = "white",
                                          place = "centre",
                                          size = 10) +
  scale_fill_viridis_d(name = "") + facet_wrap(. ~ NAME) + theme_minimal() +
  labs(title = "Arrival Time at Work",
       subtitle = "By County in Maine", caption = "Arrival Time") 
    
  
})
```


### Length of Commute

```{r}

    
renderPlot({me_traveltime %>% filter(NAME %in%input$countiesradio) %>% group_by(NAME) %>% 
    mutate(pct = estimate / (estimate[label == "Estimate!!Total"])) %>% filter(time != "Estimate!!Total") %>% ggplot(aes(
                               fill = time,
                               area = pct,
                               label = paste(time, ": ", round(pct*100, 0), "%", sep = "")))+ geom_treemap()+ geom_treemap_text(colour = "white",
                                          place = "centre",
                                          size = 10)+facet_wrap(.~NAME,scales="free")+
    scale_fill_viridis_d(name = "") + theme_minimal() +labs(title = "Travel Time to Work",
       subtitle = "By County in Maine", caption = "Travel Time") 


})
```

# Analysis

Row
-------------------------------------
    
### Travel Time to Work in Two Counties
    
```{r}

me_traveltime %>% filter(NAME %in%c("Penobscot County, Maine", "Cumberland County, Maine") & time != "Estimate!!Total") %>% ggplot(aes(
                               x = reorder(time,-estimate),
                               y = estimate,
                               fill = time
                             )) + geom_bar(position = "dodge",
                                           stat = "identity",
                                           col = "black") +facet_wrap(.~NAME,scales="free")+
    scale_fill_viridis_d(name = "") + theme_minimal() +geom_errorbar(aes(ymin=estimate-moe, ymax = estimate+moe), width=.2, position=position_dodge(.9))+
    labs(
      y = "Count",
      x = "Time",
      title = paste("Travel Time to Work")
    ) + theme(legend.position = "none") +
    theme(axis.text.x = element_text(angle = 90))


```
   
Row
-------------------------------------
    
### Means of Transportation 
    
```{r}
me_means_transport %>% filter(label != "Estimate!!Total") %>% ggplot(aes(NAME, y =
                                                                           estimate, fill = mode)) + geom_bar(position = "fill",
                                                                                                              stat = "identity",
                                                                                                              col = "black") +
  scale_fill_viridis_d(name = "") + theme_minimal() +
  labs(
    y = "Proportion",
    x = "County",
    title = "Mode of Transportation Proportion",
    subtitle = "By County in Maine", caption = "Modes of Transport"
  ) + theme(legend.position = "bottom") + scale_y_continuous(breaks = seq(0, 1, .1)) +
  theme(axis.text.x = element_text(angle = 90))

```
    
### Time of Arrival

```{r}
me_arrivetime %>% group_by(NAME) %>%
  mutate(
    pct = estimate / (estimate[label == "Estimate!!Total"]),
    NAME = str_remove(NAME, " County, Maine"),
    time = factor(
      time,
      levels = c(
        "12 00 a.m. to 4 59 a.m.",
        "5 00 a.m. to 5 29 a.m.",
        "5 30 a.m. to 5 59 a.m.",
        "6 00 a.m. to 6 29 a.m." ,
        "6 30 a.m. to 6 59 a.m."  ,
        "7 00 a.m. to 7 29 a.m." ,
        "7 30 a.m. to 7 59 a.m."  ,
        "8 00 a.m. to 8 29 a.m.",
        "8 30 a.m. to 8 59 a.m."  ,
        "9 00 a.m. to 9 59 a.m." ,
        "10 00 a.m. to 10 59 a.m.",
        "11 00 a.m. to 11 59 a.m." ,
        "12 00 p.m. to 3 59 p.m." ,
        "4 00 p.m. to 11 59 p.m."
      ),
      labels = c(
        "12-4AM",
        "5-5:30AM",
        "5:30-6AM",
        "6-6:30AM" ,
        "6:30-7AM"  ,
        "7-7:30AM" ,
        "7:30-8AM"  ,
        "8-8:30AM",
        "8:30-9AM"  ,
        "9-10AM" ,
        "10-11AM",
        "11-12AM" ,
        "12-4PM" ,
        "4-12PM"
      ),
      ordered = T
    )
  ) %>%
  filter(label != "Estimate!!Total") %>%
  ggplot(aes(
    area = pct,
    fill = time,
    label = round(pct, 2)
  )) + geom_treemap() + geom_treemap_text(colour = "white",
                                          place = "centre",
                                          size = 5) +
  scale_fill_viridis_d(name = "") + facet_wrap(. ~ NAME) + theme_minimal() +
  labs(title = "Arrival Time at Work",
       subtitle = "By County in Maine", caption = "Arrival Time") 

```


